Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-03 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

Outline of revisions:
1. generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal
2. expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation
3. changed discussion and plots to involve branch lengths and not fidelities

Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach.
Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done.
As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al.
We provide some numerical results to confirm our derivations and to provide a comparison.

%das: added
This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section.
Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
An investigation of these aspects deserves another paper.
We have added clarification of these points to the discussion:
"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners."

@@

> - line 97 below (2) define A^p

We have clarified this with an explicit definition of A^p where we introduce A, at the beginning of the section on Phylogenetic maximum likelihood.

> - l106: why is the functional form of (3) determined by inequalites;
> what are they?

We have made this more precise and directly referenced the inequalities:
"In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral states (Table~\ref{tab:farris_likelihoods}) to obtain each conditional likelihood term, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}."

@@

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3).
We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods})."

### 2018-08-03 appendix.tex
Cleaning out addressed comments.
address all comments and update responses

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
%EM perhaps a section reference for this big book?
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

%EM does this one need a different explanation now?
%das: wrote out in more detail; I realize now I maybe took some of these intermediate steps for granted. Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
%das: okay, even gorier detail---I think it's all correct now.
For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
In the case of the conditional likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
In other words, the function value for the conditional likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
Indeed,
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{2\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{2\}, \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{2\}$ and
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2,3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{1,2,3\}$, which shows the likelihood remains unchanged if $y_1$ and $y_2$ are swapped.

@@

%EM Topic sentence here please.
%das: made more explicit and introduced more gently
We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
\[
C1-C3: b_{2}^U, b_{1}^U, b_{12}^U < 1,
\]
\[
C4: -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
\]
Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}

>>>>

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
In the case of the conditional likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
In other words, the function value for the conditional likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
Indeed,
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{2\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{2\}, \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{2\}$ and
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2,3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{1,2,3\}$, which shows the likelihood remains unchanged if $y_1$ and $y_2$ are swapped.

@@

We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
\[
C1-C3: b_{2}^U, b_{1}^U, b_{12}^U < 1,
\]
\[
C4: -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
\]
Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
\caption{
    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:incons-analytic}
\end{figure}
